Packaging research artefacts with RO-Crate

### 2022-01-02 latex/ro-crate.tex
language fixes, URL for metadata in dataverse

A mapping has been produced between Research Object Crates and
Machine-actionable Data Management Plans
\cite{doi:10.4126/frl01-006423291}, implemented by the RO-Crate RDA maDMP
Mapper \cite{doi:10.5281/zenodo.3922136}. A similar mapping has been
implemented by \texttt{RO-Crate\_2\_ma-DMP} \cite{doi:10.5281/zenodo.3903463}. In
both cases, a maDMP can be converted to a RO-Crate, or vice versa. In
\cite{doi:10.4126/frl01-006423291} this functionality caters for two use cases:
\begin{enumerate}
\item[1.] Start a skeleton data management plan based on an existing RO-Crate
dataset, e.g. from an RO-Crate from WorkflowHub.
\item[2.] Instantiate an RO-Crate based on a data management plan.
\end{enumerate}

@@

The concept of a \textbf{Data Commons} for research collaboration was
originally defined as \textit{``cyber-infrastructure that co-locates data,
storage, and computing infrastructure with commonly used tools for
analysing and sharing data to create an interoperable resource for the
research community''} \cite{doi:10.1109/MCSE.2016.92}. More recently, Data
Commons has been established to mean integration of active
data-intensive research with data management and archival best
practices, along with a supporting computational infrastructure.
Furthermore, the Commons features tools and services, such as
computation clusters and storage for scalability, data repositories for
disseminating and preserving regular, but also large or sensitive
datasets, and other research assets. Multiple initiatives were
undertaken to create Data Commons on national, research, and
institutional levels. For example, the Australian Research Data
Commons (ARDC)\footnote{\url{https://ardc.edu.au}} \cite{doi:10.5334/dsj-2019-044} is a
national initiative that enables local researchers and industries to
access computing infrastructure, training, and curated datasets for
data-intensive research. NCI's Genomic Data
Commons\footnote{\url{https://gdc.cancer.gov/}} (GDC)
\cite{doi:10.1182/blood-2017-03-735654} provides the cancer research
community with access to a vast volume of genomic and clinical data.
Initiatives such as Research Data Alliance (RDA) Global Open Research
Commons\footnote{\url{https://www.rd-alliance.org/groups/global-open-research-commons-ig}}
propose standards for the implementation of Data Commons to prevent
them becoming ``data silos'' and thus, enable interoperability from one
Data Commons to another.

@@

Particularly relevant to this article is the second objective of the
Harvard Data Commons, which aims to support the deposit of research
artefacts to Harvard Dataverse with sufficient information in the
metadata to allow their future reuse (Fig.~\ref{fig:hdc}). To support
the incorporation of data, code, and other artefacts from various
institutional infrastructures, Harvard Data Commons is currently
working on RO-Crate adaptation. The RO-Crate metadata provides the
necessary structure to make all research artefacts FAIR. The Dataverse
software already has extensive support for metadata, including the Data
Documentation Initiative (DDI), Dublin Core, DataCite, and Schema.org.
Incorporating RO-Crate, which has the flexibility to describe a wide
range of research resources, will facilitate their seamless transition
from one infrastructure to the other within the Harvard Data Commons.

@@

Several developments for Research Objects improved on this situation,
such as ROHub used by Earth Sciences
\cite{doi:10.1016/j.future.2019.03.046}, which provides a interactive
user-interface for making Research Objects, along with Research Object
Bundle \cite{doi:10.5281/zenodo.12586} (RO Bundle), which is a ZIP-archive
embedding data files and a JSON-LD serialization of the manifest with
mappings for a limited set of terms. RO Bundle was also used for
storing detailed workflow run provenance (TavernaPROV
\cite{doi:10.5281/zenodo.51314}).

@@

The use of computational workflows, typically combining a chain of
tools in an analytical pipeline, has gained prominence in particular in
the life sciences. Workflows might be used primarily to improve
computational scalability, as well as to also assist in making computed
data results FAIR \cite{doi:10.1162/dint_a_00033}, for instance by
improving reproducibility \cite{doi:10.1016/j.future.2017.01.012}, but also
because programmatic data usage help propagate their metadata and
provenance \cite{doi:10.1002/cpe.1228}. At the same time, workflows raise
additional FAIR challenges, since they can be considered important
research artefacts themselves. This viewpoint poses the problem of
capturing and explaining the computational methods of a pipeline in
sufficient machine-readable detail \cite{doi:10.3233/DS-190026}.

>>>>

A mapping has been produced between Research Object Crates and
Machine-actionable Data Management Plans
\cite{doi:10.4126/frl01-006423291}, implemented by the RO-Crate RDA maDMP
Mapper \cite{doi:10.5281/zenodo.3922136}. A similar mapping has been
implemented by \texttt{RO-Crate\_2\_ma-DMP} \cite{doi:10.5281/zenodo.3903463}. In
both cases, a maDMP can be converted to a RO-Crate, or vice versa. In
\cite{doi:10.4126/frl01-006423291} this functionality caters for two use cases:
\begin{enumerate}
\item[1.] Start a skeleton data management plan based on an existing RO-Crate
dataset, e.g. an RO-Crate from WorkflowHub.
\item[2.] Instantiate an RO-Crate based on a data management plan.
\end{enumerate}

@@

The concept of a \textbf{Data Commons} for research collaboration was
originally defined as \textit{``cyber-infrastructure that co-locates data,
storage, and computing infrastructure with commonly used tools for
analysing and sharing data to create an interoperable resource for the
research community''} \cite{doi:10.1109/MCSE.2016.92}. More recently, Data
Commons has been established to mean integration of active
data-intensive research with data management and archival best
practices, along with a supporting computational infrastructure.
Furthermore, the Commons features tools and services, such as
computation clusters and storage for scalability, data repositories for
disseminating and preserving regular, but also large or sensitive
datasets, and other research assets. Multiple initiatives were
undertaken to create Data Commons on national, research, and
institutional levels. For example, the Australian Research Data
Commons (ARDC)\footnote{\url{https://ardc.edu.au/}} \cite{doi:10.5334/dsj-2019-044} is a
national initiative that enables local researchers and industries to
access computing infrastructure, training, and curated datasets for
data-intensive research. NCI's Genomic Data
Commons\footnote{\url{https://gdc.cancer.gov/}} (GDC)
\cite{doi:10.1182/blood-2017-03-735654} provides the cancer research
community with access to a vast volume of genomic and clinical data.
Initiatives such as Research Data Alliance (RDA) Global Open Research
Commons\footnote{\url{https://www.rd-alliance.org/groups/global-open-research-commons-ig}}
propose standards for the implementation of Data Commons to prevent
them becoming ``data silos'' and thus, enable interoperability from one
Data Commons to another.

@@

Particularly relevant to this article is the second objective of the
Harvard Data Commons, which aims to support the deposit of research
artefacts to Harvard Dataverse with sufficient information in the
metadata to allow their future reuse (Fig.~\ref{fig:hdc}). To support
the incorporation of data, code, and other artefacts from various
institutional infrastructures, Harvard Data Commons is currently
working on RO-Crate adaptation. The RO-Crate metadata provides the
necessary structure to make all research artefacts FAIR. The Dataverse
software already has extensive support for metadata\footnote{\url{https://guides.dataverse.org/en/latest/user/appendix.html}}, including the Data
Documentation Initiative (DDI), Dublin Core, DataCite, and Schema.org.
Incorporating RO-Crate, which has the flexibility to describe a wide
range of research resources, will facilitate their seamless transition
from one infrastructure to the other within the Harvard Data Commons.

@@

Several developments for Research Objects improved on this situation,
such as ROHub used by Earth Sciences
\cite{doi:10.1016/j.future.2019.03.046}, which provides a 
user-interface for making Research Objects, along with Research Object
Bundle \cite{doi:10.5281/zenodo.12586} (RO Bundle), which is a ZIP-archive
embedding data files and a JSON-LD serialization of the manifest with
mappings for a limited set of terms. RO Bundle was also used for
storing detailed workflow run provenance (TavernaPROV
\cite{doi:10.5281/zenodo.51314}).

@@

The use of computational workflows, typically combining a chain of
tools in an analytical pipeline, has gained prominence in particular in
the life sciences. Workflows might be used primarily to improve
computational scalability, as well as to assist in making computed
data results FAIR \cite{doi:10.1162/dint_a_00033}, for instance by
improving reproducibility \cite{doi:10.1016/j.future.2017.01.012}, but also
because programmatic data usage help propagate their metadata and
provenance \cite{doi:10.1002/cpe.1228}. At the same time, workflows raise
additional FAIR challenges, since they can be considered important
research artefacts themselves. This viewpoint poses the problem of
capturing and explaining the computational methods of a pipeline in
sufficient machine-readable detail \cite{doi:10.3233/DS-190026}.
